# Phylogenetic Independent Contrasts {#pic}


Let's make a digression to look at Phylogenetic Independent Contrasts (PIC). PIC were the first comparative approach proposed to deal with phylogenetic non independence [@felsenstein1985phylogenies]. Although they are less flexible than PGLS, they give the same results. Let's see how they can be used.

Phylogenetic independent contrast are estimated one trait at a time. They essentially transform the observed trait in contrasts that are not correlated with the phylogeny. This can be done in R using the `pic` function of the `ape` package.

```{r estimating_contrast}
# Estimate PIC for shade tolerance 
Shade.pic <- pic(seedplantsdata$Shade, phy=seedplantstree)
# Estimate PIC for Wood density 
Wd.pic <- pic(seedplantsdata$Wd, phy=seedplantstree)
```

Once this is done, the only thing to do is to fit a regression between these contrast. Note that it is important that the intercept is fixed to 0 in the model. This is done by adding `- 1` to the right of the formula.

```{r fitting_pic}
# Estimate PIC for shade tolerance
pic.results <- lm(Shade.pic ~ Wd.pic - 1)
summary(pic.results)
```

You can see that the slope estimate, `4.361`, it identical to the slope estimate obtained with PGLS. Same thing for the p-value. The main retriction with PIC is that you are limited in always comparing two variables. Much more flexibility is possible with PGLS.


# Relaxing the assumption that residuals need to be perfectly phylogenetically correlated

Phylogenetic Generalized Least Squares assume that the residuals are perfectly phylogenetically correlated. This is relatively constraining because it means that other sources of errors that are not phylogenetically correlated are not allowed by the model. Moreover, if these exist, they can bias the results of the PGLS [@revell2010phylogenetic].

There are ways to relax this assumption, and one of this is to use a type of correlation structure that allows to relax this assumption.


## Theory: Pagel's correlation structure

When controling for phylogenetic relationships with phylogenetic generalized least squares, we assume that the residuals are perfectly correlated according to the correlation structure. In practice, it might not be always the case and it is difficult to really know how important it is to control for the phylogenetic relationship in a specific case. For instance, for a given study, the correlation in the residuals might not be highly phylogenetically correlated.

This is possible to account for this using the $\lambda$ model of Pagel [@pagel1999inferring]. The idea is to multiply the off-diagonal of the correlation matrix (essentially the branch lengths of the phylogeny) by a parameter $\lambda$, but not the diagonal values. This essentially leads to a modification of branch lengths of the phylogeny. A $\lambda$ value near zero gives very shorts internal branches and long tip branches. This, in effect, reduces the phylogenetic correlations (the effect of the phylogeny is reduced). At the opposite, if $\lambda$ is close to 1, then the modified phylogeny resembles the true phylogeny. Indeed, the parameter $\lambda$ is often interpreted as a parameter of phylogenetic signal; as such, a greater $\lambda$ value implies a stronger phylogenetic signal.

The following figure shows how different lambda values affect the shape of the Quebec trees phylogeny.

```{r "alpha example", echo=FALSE, fig.align='center', message=FALSE}
require(geiger)
op <- par(mfrow=c(1,3))
plot(rescale(seedplantstree,model="lambda",0.1),main=expression("lambda=0.1"))
plot(rescale(seedplantstree,model="lambda",0.5),main=expression("lambda=0.5"))
plot(rescale(seedplantstree,model="lambda",1),main=expression("lambda=1"))
par(op)
```

You can see that with small values of lambda, the weight given to the shared history (the phylogeny) are greatly reduced. The long terminal branches somewhat indicates that there could be a lot more variation in the residuals that are independent of the other species. This variation could be due to other factors that are included in the estimates of each species but that are independent of the phylogeny (such as measurement errors for instance).

## Practicals

Pagel's $\lambda$ model can be used in PGLS using the `corPagel` correlation structure. The usage of this correlation structure is similar to that of the `corBrownian` structure, except that you need to provide a starting parameter value for $\lambda$.

```{r "corPagel example"}
# Get the correlation structure
pagel.corr <- corPagel(0.3, phy=seedplantstree, fixed=FALSE, form=~rownames())
```

>In some version of R, the `pagel.corr` structure seems to be giving problems. If you get an error when fitting it with the `gls` function below, then remove the option `form=~rownames()` from the `corPagel` function. Note that if you do this, you need to be very sure that the order of the tips of the tree (seedplantstree$tip.label) is the same as the row of the seedplantsdata table.

The value given to `corPagel` is the starting value for the $\lambda$ parameter. Also, note that the option `fixed=` is set to `FALSE` This means that the $\lambda$ parameter will be optimized using generalized least squares. If it was set to `TRUE`, then the model would be fitted with the starting parameter, here `0.3`.

Let's now fit the PGLS with this correlation structure.

```{r "Fitting PGLS with corPagel"}
# PGLS with coraPagel
shade.pgls2 <- gls(Shade ~ Wd, data = seedplantsdata, correlation=pagel.corr)
summary(shade.pgls2)
```

You can see that gls has estimated the $\lambda$ parameter, which is 0.958 here. Because the estimated $\lambda$ is very close to 1, we can conclude that residuals of the model were strongly phylogenetically correlated. This, in turns, thus confirms the importance of using a PGLS with this model. If the $\lambda$ estimated would have been close to 0, it would have suggested that the PGLS is not necessary. Note, however, that using this approach assures you to never obtained a biased statistical result. Actually, I **strongly recommend** that you always use this correlation structure in your statistical analyses.


## Challenge 4

Try to fit a PGLS with a Pagel correlation structure when regressing Shade tolerance on seed mass. Are the residuals as phylogenetically correlated than in the previous regression with wood density?

```{r "Challenge 4", warning=FALSE, echo=FALSE, eval=FALSE}
# Fit a PGLS with the gls function
Sm.pgls2 <- gls(Shade ~ Sm, data = seedplantsdata, correlation=pagel.corr)
# Get the results
summary(Sm.pgls2)
```


## Other correlation structures (or evolutionary models)

The correlation structures available in the package `ape` offer other alternatives for the assumed model of character evolution. For instance, the `corMartins` correlation structure models selection using the Ornstein-Uhlenbeck (or Hansen) model with parameter $\alpha$ that determines the strength of the selection. Also, `corBlomberg` models accelerating or decelerating Brownian evolution, that is, the evolutionary rate of the Brownian motion is either accelerating or decelerating with time with this model. It is possible to do model comparisons to decide which model best fit the residual variation.
