# Phylogenetic ANOVA


So far, we have only analysed continuous quantitative characters. But it is also possible to perform an ANOVA with PGLS.

The great thing with PGLS as implemented with the `gls` function is that it can easily be adapted to testing many different types of models. To give just one example here, it is easy to implement a phylogenetic ANOVA in R. Indeed, you just need to give `gls` a categorical trait as independent variable.

Because there is no categorical variable in the plant functional trait dataset, we will create one by dividing the wood density category in two categories, light and dense wood.

```{r "make categorical variable"}
# Make categorical variable
seedplantsdata$Wd.cat<-cut(seedplantsdata$Wd,breaks=2,labels=c("light","dense"))
# Look at the result
seedplantsdata$Wd.cat
```

We can now fit a phylogenetic ANOVA.

```{r "Phylogenetic ANOVA"}
# Phylogenetic ANOVA
shade.pgls3 <- gls(Shade ~ Wd.cat, data = seedplantsdata, correlation=pagel.corr)
summary(shade.pgls3)
```

You can see that the wood density, even when transformed in a categorical variable, has a significant effect on shade tolerance.
