# Solutions to the challenges {#solutions}

## Challenge 1

In the `seedplantsdata` data frame, there were many different traits. Try to fit a regression of tree shade tolerance (`shade`) on the seed mass (`Sm`). In other words, test if shade tolerance can be explained by the seed mass of the trees. Then, try to see if the residuals are phylogenetically correlated.

```{r "Challenge 1 - solution", warning=FALSE, echo=TRUE, eval=TRUE}
# Fit a linear model using Ordinary Least Squares (OLS)
Sm.lm <- lm(Shade ~ Sm, data = seedplantsdata)
# Get the results
summary(Sm.lm)
# Extract the residuals
Sm.res <- residuals(Sm.lm)
# Plot the residuals beside the phylogeny
op <- par(mar=c(1,1,1,1))
plot(seedplantstree,type="p",TRUE,label.offset=0.01,cex=0.5,no.margin=FALSE)
tiplabels(pch=21,bg=cols[ifelse(Sm.res>0,1,2)],col="black",cex=abs(Sm.res),adj=0.505)
legend("topleft",legend=c("-2","-1","0","1","2"),pch=21,
       pt.bg=cols[c(1,1,1,2,2)],bty="n",
       text.col="gray32",cex=0.8,pt.cex=c(2,1,0.1,1,2))
par(op)
```


## Challenge 2

Can you get the covariance matrix and the correlation matrix for the seed plants phylogenetic tree from the example above (`seedplantstree`)?

```{r "Challenge 2 - solution", echo=TRUE, eval=TRUE}
# Covariance matrix
seedplants.cov <- vcv(seedplantstree,corr=FALSE)
# Check the first few lines of the matrix
head(round(seedplants.cov,3))
# Correlation matrix
seedplants.cor <- vcv(seedplantstree,corr=TRUE)
# Check the first few lines of the matrix
head(round(seedplants.cor,3))
```


## Challenge 3

Fit a PGLS model to see whether the seed mass (`Sm`) explains shade tolerance (`Shade`) with the seedplantdataset. How does it compare to the results from the standard regression.

```{r "Challenge 3 - solution", warning=FALSE, echo=TRUE, eval=TRUE}
# Fit a PGLS with the gls function
Sm.pgls <- gls(Shade ~ Sm, data = seedplantsdata, correlation=bm.corr)
# Get the results
summary(Sm.pgls)
# Extract the residuals corrected by the correlation structure
Sm.pgls.res <- residuals(Sm.pgls,type="normalized")
# Plot the residuals beside the phylogeny
op <- par(mar=c(1,1,1,1))
plot(seedplantstree,type="p",TRUE,label.offset=0.01,cex=0.5,no.margin=FALSE)
tiplabels(pch=21,bg=cols[ifelse(Sm.pgls.res>0,1,2)],col="black",cex=abs(Sm.pgls.res),adj=0.505)
legend("topleft",legend=c("-2","-1","0","1","2"),pch=21,
       pt.bg=cols[c(1,1,1,2,2)],bty="n",
       text.col="gray32",cex=0.8,pt.cex=c(2,1,0.1,1,2))
par(op)
```


## Challenge 4

Try to fit a PGLS with a Pagel correlation structure when regressing Shade tolerance on seed mass. Are the residuals as phylogenetically correlated than in the previous regression with wood density?

```{r "Challenge 4 - solution", warning=FALSE, echo=TRUE, eval=TRUE}
# Fit a PGLS with the gls function
Sm.pgls2 <- gls(Shade ~ Sm, data = seedplantsdata, correlation=pagel.corr)
# Get the results
summary(Sm.pgls2)
```

